Automatic Differentiation Accelerates Inference for Partially Observed Markov Processes

### 2024-10-17 paper/ms.tex
Updates from Overleaf

\cite{scibior21} showed that the estimators derived by \cite{poyiadjis11} can be attained with standard AD software using an algorithmic procedure, called a {\it stop-gradient}, which allows selected expressions to be evaluated but not differentiated.
However, the algorithm of \cite{scibior21} that can be modified to have the plug-and-play property recovers an estimator with high Monte Carlo variability, which led the authors to reject this algorithm in favor of an approach without the plug-and-play property and with poorer scalability.

@@

We therefore see that the second stage of IFAD converges linearly to the MLE if (1) the log-likelihood surface is well-conditioned in a neighborhood of the MLE and (2) the first stage of IFAD successfully reaches a (high-probability) basin of attraction of the MLE.
Within a neighborhood of the MLE that grows as more data become available \citep{ning21}, local quadratic approximations such as local asymptotic normality apply under general conditions \citep{lecam00}.
This suggests that the linear convergence may occur reasonably often in practice.

@@

\begin{algorithm}[H]
	\caption{IFAD}
    \label{alg:ifad}
	    \textbf{Input:} Number of particles $J$, timesteps $N$, IF2 cooling schedule $\eta_m$, MOP-$\alpha$ discounting parameter $\alpha$, $\theta_0$, $m=0.$\newline
        Run IF2 until initial "convergence" under cooling schedule $\eta_m$ to obtain $\{\Theta_j, j=1,...,J\}$, set $\theta_m := \frac{1}{J}\sum_{j=1}^J \Theta_j.$\newline
		\textbf{While} procedure not converged: \newline
		\hspace*{4mm} Run Algorithm \ref{alg:mop} to obtain $\hat\loglik(\theta_m).$ \newline
		\hspace*{4mm} $g(\theta_m) := \nabla_{\theta}(-\hat\loglik(\theta_m))$, $H(\theta_m)$ s.t. $\lambda_{\min}(H(\theta_m)) \geq c$. \newline
		\hspace*{4mm} Update $\theta_{m+1} := \theta_m - \eta (H(\theta_m))^{-1} g(\theta_m)$, $m:=m+1.$ \newline
		\textbf{Return} $\hat{\theta} := \theta_m.$
\end{algorithm}

>>>>

\cite{scibior21} showed that the estimators derived by \cite{poyiadjis11} can be attained with standard AD software using an algorithmic procedure, called a {\it stop-gradient}, which allows selected expressions to be evaluated but not differentiated.
However, the algorithm of \cite{scibior21} that can be modified to have the plug-and-play property recovers an estimator with high Monte Carlo variability, which led the authors to reject this algorithm in favor of an approach without the plug-and-play property and with inferior scalability.

@@

We therefore see that the second stage of IFAD converges linearly to the MLE if (1) the log-likelihood surface is well-conditioned in a neighborhood of the MLE and (2) the first stage of IFAD successfully reaches a (high-probability) basin of attraction of the MLE.
Within a neighborhood of the MLE that grows as more data become available \citep{ning21}, local quadratic approximations such as local asymptotic normality apply under general conditions \citep{lecam00}.
This suggests that the linear convergence may occur often in practice.

@@

\begin{algorithm}[H]
	\caption{IFAD}
    \label{alg:ifad}
	    \textbf{Input:} Number of particles $J$, timesteps $N$, IF2 cooling schedule $\eta_m$, MOP-$\alpha$ discounting parameter $\alpha$, $\theta_0$, $m=0.$\newline
        Run IF2 until initial ``convergence'' under cooling schedule $\eta_m$ to obtain $\{\Theta_j, j=1,...,J\}$, set $\theta_m := \frac{1}{J}\sum_{j=1}^J \Theta_j.$\newline
		\textbf{While} procedure not converged: \newline
		\hspace*{4mm} Run Algorithm \ref{alg:mop} to obtain $\hat\loglik(\theta_m).$ \newline
		\hspace*{4mm} $g(\theta_m) := \nabla_{\theta}(-\hat\loglik(\theta_m))$, $H(\theta_m)$ s.t. $\lambda_{\min}(H(\theta_m)) \geq c$. \newline
		\hspace*{4mm} Update $\theta_{m+1} := \theta_m - \eta (H(\theta_m))^{-1} g(\theta_m)$, $m:=m+1.$ \newline
		\textbf{Return} $\hat{\theta} := \theta_m.$
\end{algorithm}
